---
title: "Plots and analyses"
author: "APB"
date: "5/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(geiger)
library(phytools)
library(nlme)
library(caper)
library(gbm)
library(ggplot2)
library(sf)
library(picante)
library(raster)
library(scico)
library(spdep)
library(gbm)
library(caret)
library(ggpubr)
```

```{r load_data}
#all trait data per grid cell
trait_npp_data<-readRDS("trait_npp_data.rds") 
#phylogeny
tree<-readRDS("SA_tree.rds")
#trait data per tree tip
tree_data<-readRDS("tree_data.rds")
#put tree data in tree tip.label order
tree_data_ordered<-tree_data[match(tree$tip.label, tree_data$matrix_name),]
```

#Is there phylogenetic signal in tree traits?
```{r test_phylo_signal, echo = TRUE}
##create named vectors for traits of interest
height<-tree_data_ordered$height
names(height)<-tree_data_ordered$matrix_name
ratio<-tree_data_ordered$ratio
names(ratio)<-tree_data_ordered$matrix_name
leafarea<-tree_data_ordered$leaf_area
names(leafarea)<-tree_data_ordered$matrix_name

##check for phylogenetic signal

#phylosig in the package phytools
phylosig(tree, height, method = "lambda")   #0.848
phylosig(tree, ratio, method = "lambda")    #0.818
phylosig(tree, leafarea, method = "lambda") #0.684

#gls() from the 'nlme' package comparing a null model to a model that includes the only the tree
##remove NAs and exotics
tree_data_ordered %>%
  filter(height != "Exotic") %>%
  filter(height != "NA") %>% 
  filter(ratio != "NA")-> corr_data
##make a tree to match
corr_tree<-drop.tip(tree, tree$tip.label[which(!(tree$tip.label %in% corr_data$matrix_name))])

##make height models
height_null<-gls(log(height) ~ 1, corr_data)
height_tree<-gls(log(height) ~ 1, corr_data, correlation = corPagel(0.5, corr_tree)) # lambda = 0.771
##compare models
anova(height_null, height_tree) #model with phylogeny is better (p < 0.0001)

##make leaf shape models
ratio_null<-gls(log(ratio) ~ 1, corr_data)
ratio_tree<-gls(log(ratio) ~ 1, corr_data, correlation = corPagel(0.5, corr_tree)) # lambda = 0.802
##compare models
anova(ratio_null, ratio_tree) #model with phylogeny is better (p < 0.0001)

##make leaf area models
leafarea_null<-gls(log(leaf_area) ~ 1, corr_data)
leafarea_tree<-gls(log(leaf_area) ~ 1, corr_data, correlation = corPagel(0.5, corr_tree)) # lambda = 0.921
##compare models
anova(leafarea_null, leafarea_tree) #model with phylogeny is better (p < 0.0001)
```

There is phylogenetic signal in tree height, leaf length:width, and leaf area. What does it look like?
```{r plot_traits, echo = FALSE}
height_values<-log(corr_data$height)
names(height_values)<-corr_tree$tip.label
contMap(corr_tree,height_values, fsize = 0.2, lwd = 0.4, type= "fan")

shape_values<-log(corr_data$ratio)
names(shape_values)<-corr_tree$tip.label
contMap(corr_tree, shape_values, fsize = 0.2, lwd = 0.4, type= "fan") 
```


Are there correlations between these traits? Are there phylogenetic correlations between these traits? 
```{r correlation, echo = TRUE}
#Is there a correlation between tree height and leaf shape? 
cor.test(log(corr_data$height),log(corr_data$ratio)) # yes -.15
summary(lm(log(height) ~ log(ratio), corr_data))

fig2a<-ggplot(corr_data, mapping=aes(x= log(ratio), y = log(height)))+
  geom_point(pch = 1, size = 2, alpha = 0.5, color = "#ffcc33", )+
  geom_smooth(method = "lm", color = "black", alpha = 0.5, linewidth = .5)+
  labs(y = "log (Tree Height)", x = " ")+
  theme_classic()+
  theme(text = element_text(size=12),
        axis.title = element_text(face = "bold", size =12),
        strip.background =element_rect(fill="transparent", color="black", linewidth = 1),
        strip.text = element_text(face = "bold",size=12, color="black"),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10),
        panel.background = element_rect(fill="transparent", color =NA),
        #panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank(),
        panel.border = element_rect(fill="transparent", color="black", linewidth = 1))+
  stat_cor(method = "pearson", label.x = 2.5, label.y =4)

#Is there a phylogenetic correlation? (pgls() from the caper package)
## make a data frame
dat<-data.frame(taxa=corr_tree$tip.label,
                log_height=log(corr_data$height), 
                log_leaf_shape=log(corr_data$ratio),
                log_leaf_area=log(corr_data$leaf_area))
##make a comparative data frame required for pgls
cdat<-comparative.data(data = dat, phy = corr_tree, names.col="taxa")

#Is there a phylogenetic correlation between tree height and leaf shape? 
summary(height_pgls<-pgls(log_height ~ log_leaf_shape, cdat, lambda = "ML")) #NO multiple R2 = 0.00082
summary(tree_height_gls<-gls(log(height) ~ log(ratio), corr_data, correlation = corPagel(0.5, corr_tree))) # r = -0.108, p = 0.348
plot(tree_height_gls)
abline(a = coef(tree_height_gls)[1], b = coef(tree_height_gls)[2])

fig2b<-corr_data %>%
  ggplot()+
  geom_point(mapping=aes(x=log(ratio), y=log(height)), pch = 1, size = 2, alpha = 0.5, color = "#ffcc33")+
  #geom_abline(intercept = coef(height_pgls)[1], slope = coef(height_pgls)[2],color = "black")+
  #geom_segment(aes(x= -1.5, xend = 3.8, y = coef(height_pgls)[1] + coef(height_pgls)[2], yend = coef(height_pgls)[1] + coef(height_pgls)[2]*3.8))+
  annotate('segment', x= -1.5, xend = 3.8, y = coef(height_pgls)[1] + coef(height_pgls)[2], yend = coef(height_pgls)[1] + coef(height_pgls)[2]*3.8, linewidth = 0.5)+
  labs(y = "log (Tree Height)", x = " ")+
  annotate(geom= 'text', x=3.2, y=4, label=expression(paste(italic("R = "),"-0.03, ",italic("p = "),"3.5e-01")), parse = TRUE)+
  theme_classic()+
  theme(text = element_text(size=12),
        axis.title = element_text(face = "bold", size =12),
        strip.background =element_rect(fill="transparent", color="black", linewidth = 1),
        strip.text = element_text(face = "bold",size=12, color="black"),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10),
        panel.background = element_rect(fill="transparent", color =NA),
        #panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank(),
        panel.border = element_rect(fill="transparent", color="black", linewidth = 1))
 
  

#Is there a correlation between tree height and leaf area?
cor.test(log(corr_data$height), log(corr_data$leaf_area)) #Yes 0.268
summary(lm(log(height) ~ log(leaf_area), corr_data))

#Is there a phylogenetic correlation?
summary(pgls(log_height ~ log_leaf_area, cdat, lambda = "ML")) #YES 0.34
summary(gls(log(height) ~ log(leaf_area), corr_data, correlation = corPagel(0.5, corr_tree)))

#Is there a correlation between leaf shape and leaf area?
cor.test(log(corr_data$leaf_area), log(corr_data$ratio)) #no 0.064
plot(log(leaf_area) ~ log(ratio), corr_data)
```


#Grid level analyses
```{r correlations, echo = TRUE}
#Is there a correlation between leaf shape and NPP? YES
summary(lm(log(leafshape) ~ NPP, data = trait_npp_data))
cor.test(log(trait_npp_data$leafshape), trait_npp_data$NPP) #r = -0.578
plot(log(leafshape) ~ NPP, data = trait_npp_data) #rounder leaves in areas with higher NPP

#Is there a correlation between leaf area and NPP? YES
summary(lm(log(leafarea) ~ NPP, data = trait_npp_data))
cor.test(log(trait_npp_data$leafarea), trait_npp_data$NPP) #r = 0.58
plot(log(leafarea) ~ NPP, data = trait_npp_data) #leaves with larger surface areas in grids with higher NPP

#Is there a correlation between tree height and species richness? YES
summary(lm(log(treeheight) ~ log(ntaxa), data = trait_npp_data)) 
cor.test(log(trait_npp_data$treeheight), log(trait_npp_data$ntaxa)) #r = 0.821
plot(log(treeheight) ~ log(ntaxa), data = trait_npp_data) #taller trees in areas with more species 


#Is there a correlation between leaf shape and tree height? #YES
summary(lm(log(treeheight) ~ log(leafshape), data = trait_npp_data)) 
cor.test(log(trait_npp_data$treeheight), log(trait_npp_data$leafshape)) #r= -0.764
plot(log(trait_npp_data$treeheight) ~ log(trait_npp_data$leafshape))

fig2c<-ggplot(trait_npp_data, mapping=aes(x= log(leafshape), y = log(treeheight)))+
  geom_point(pch = 1, size = 2, alpha = 0.2, color = "#ffcc33")+
  geom_smooth(method = "lm",  color = "black", alpha = 0.5, linewidth = .5)+
  labs(y = "log (Tree Height)", x = "log (Leaf Shape)")+
  theme_classic()+
  theme(text = element_text(size=12),
        axis.title = element_text(face = "bold", size =12),
        strip.background =element_rect(fill="transparent", color="black", linewidth = 1),
        strip.text = element_text(face = "bold",size=12, color="black"),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10),
        panel.background = element_rect(fill="transparent", color =NA),
        #panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank(),
        panel.border = element_rect(fill="transparent", color="black", linewidth = 1))+
  stat_cor(method = "pearson", label.x = 1.5, label.y =3)
```
```{r figure2, echo = FALSE}
library(patchwork)

fig2a / fig2b / fig2c + plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(face = 'bold'))

ggsave("figure2.png", width = 8, height = 12)
```


#Model of grid cells
mpd is the response for the outcomes of the mayfield and levine 2010
 
moran's I on the residuals(autocorrelation in space)

```{r models, include = TRUE}
#linear model predicting observed mpd 
full_model_obs<-lm(mpd.obs ~ log(treeheight) + log(leafshape) + log(ntaxa) + NPP, data = trait_npp_data)
summary(full_model_obs)
plot(full_model_obs)

#map the residulals of the full model
plot(full_model_obs$residuals)

#linear model predicting SES of mpd
full_model_ses<-lm(mpd.obs.z ~ log(treeheight) + log(leafshape) + log(ntaxa) + NPP, data = trait_npp_data)
summary(full_model_ses)
plot(full_model_ses)

#map the residuals
plot(full_model_ses$residuals)

###calculate Moran's I for the residuals

#determine neighbours
neighbours <- poly2nb(trait_npp_data)

#make weights matrix
listw <- nb2listw(neighbours)

#calculate Moran's I
#observed mpd
globalMoran_obs <- moran.test(full_model_obs$residuals, listw)
globalMoran_obs #0.73  the residuals of the model are positively autocorrelated
#SES mpd
globalMoran_ses <- moran.test(full_model_ses$residuals, listw)
globalMoran_ses #0.84  the residuals of the model are positively autocorrelated
#plot the correlations
moran.plot(full_model_obs$residuals, listw = nb2listw(neighbours, style = "W"))
moran.plot(full_model_ses$residuals, listw = nb2listw(neighbours, style = "W"))


###
model_lat<-lm(mpd.obs.z ~ log(treeheight) + log(leafshape) + log(ntaxa) + NPP + lat + long, data = trait_npp_data)
summary(model_lat)
plot(model_lat)

plot(model_lat$residuals)

#check for collinearity
car::vif(model_lat)

#calculate Moran's I
latMoran<-moran.test(model_lat$residuals, listw)
latMoran #still has a high spatial autocorrelation 0.73

#plot the correlation
moran.plot(model_lat$residuals, listw = nb2listw(neighbours, style = "W"))
```

 
#boosted regression model (mpd ~ everything) 
```{r boosted_regression, echo =TRUE}
#clean data
bgr_data<-trait_npp_data[,c(1,2,7,12,16,17,19)]
bgr_data<- bgr_data %>% st_drop_geometry()
bgr_data$log_treeheight<-log(bgr_data$treeheight) #log the height
bgr_data$log_leafshape<-log(bgr_data$leafshape) #log the leaf shape
bgr_data$log_ntaxa<-log(bgr_data$ntaxa) #log the species richness
bgr_data2<-bgr_data[,c(1,2,4,7,8,9,10)] # only include the traits for the model

#save dataframe
saveRDS(bgr_data2, "bgr_data2.rds")

# for reproducibility
set.seed(456)
#create training (70%) and test (30%) datasets 
library(rsample)
d_split<-initial_split(bgr_data2, prop = .7)
bgr_train<-training(d_split)
bgr_test<-testing(d_split)


# train GBM model
gbm.fit <- gbm(
  formula = mpd.obs.z ~ .,
  distribution = "gaussian",
  data = bgr_train,
  n.trees = 10000,
  interaction.depth = 1,
  shrinkage = 0.001,
  cv.folds = 5,
  n.cores = NULL, # will use all cores by default
  verbose = FALSE
  ) 

print(gbm.fit)
# get MSE (mean square error)
sqrt(min(gbm.fit$cv.error))
## [1] 0.4695194 the model is off by ~.5 units of mpd.z

# plot loss function as a result of n trees added to the ensemble
gbm.perf(gbm.fit, method = "cv")

# train GBM model (increase depth and decrease trees)
gbm.fit2 <- gbm(
  formula = mpd.obs.z ~ .,
  distribution = "gaussian",
  data = bgr_train,
  n.trees = 5000,
  interaction.depth = 3, #increase the depth of each tree
  shrinkage = 0.1,       #increase the learning rate to take larger steps down the gradient descent
  cv.folds = 5,
  n.cores = NULL,
  verbose = FALSE
  )
print(gbm.fit2)

#save the model for easy use
saveRDS(gbm.fit2, "gbm.fit2.rds")
#load best fitting model
#gbm.fit2<-readRDS("gbm.fit2.rds")

# find index for n trees with minimum CV error
min_MSE <- which.min(gbm.fit2$cv.error)

# get MSE and compute RMSE (root mean square error)
sqrt(gbm.fit2$cv.error[min_MSE])
## [1]  0.1050312 this is a better fit

# plot loss function as a result of n trees added to the ensemble
gbm.perf(gbm.fit2, method = "cv")

#visualize variable importance
par(mar = c(5, 8, 1, 1))
summary(
  gbm.fit2, 
  cBars = 6, #the number of variables to show (in order of influence)
  method = relative.influence, # variables with the largest average decrease in MSE are considered most important (also can use permutation.test.gbm)
  las = 2
  )
```

```{r partial_dependence_plots, echo = TRUE}
#How does the response variable change based on the variables?
library(pdp)

#Look at Leaf Shape
##average marginal effect on mpd.obs.z
gbm.fit2 %>%
  pdp::partial(pred.var = "log_leafshape", n.trees = gbm.fit2$n.trees, grid.resolution = 100) %>%
  autoplot(rug = TRUE, train = bgr_train) +
  ylab("Standard Effect Size of MPD")

#ICE (individual conditional expectation) plot
##change in predicted response variable for each observation as each predictor variable is varied
gbm.fit2 %>%
  partial(
    pred.var = "log_leafshape", 
    n.trees = gbm.fit2$n.trees, 
    grid.resolution = 100,
    ice = TRUE
    ) %>%
  autoplot(rug = TRUE, train = bgr_train, alpha = .1) +
  ggtitle("Non-centered")

#Look at Species Richness
##average marginal effect on mpd.obs.z
gbm.fit2 %>%
  pdp::partial(pred.var = "log_ntaxa", n.trees = gbm.fit2$n.trees, grid.resolution = 100) %>%
  autoplot(rug = TRUE, train = bgr_train) +
  ylab("MPD.Z")

#ICE plot
##change in predicted response variable for each observation as each predictor variable is varied
gbm.fit2 %>%
  partial(
    pred.var = "log_ntaxa", 
    n.trees = gbm.fit2$n.trees, 
    grid.resolution = 100,
    ice = TRUE
    ) %>%
  autoplot(rug = TRUE, train = bgr_train, alpha = .1) 

#Look at latitude
##average marginal effect on mpd.obs.z
gbm.fit2 %>%
  pdp::partial(pred.var = "lat", n.trees = gbm.fit2$n.trees, grid.resolution = 100) %>%
  autoplot(rug = TRUE, train = bgr_train) +
  ylab("MPD.Z")

#ICE plot
gbm.fit2 %>%
  partial(
    pred.var = "lat", 
    n.trees = gbm.fit2$n.trees, 
    grid.resolution = 100,
    ice = TRUE
    ) %>%
  autoplot(rug = TRUE, train = bgr_train, alpha = .1) 

#Look at Tree height
##average marginal effect on mpd.obs.z
gbm.fit2 %>%
  pdp::partial(pred.var = "log_treeheight", n.trees = gbm.fit2$n.trees, grid.resolution = 100) %>%
  autoplot(rug = TRUE, train = bgr_train) +
  ylab("MPD.Z")

#ICE plot
gbm.fit2 %>%
  partial(
    pred.var = "log_treeheight", 
    n.trees = gbm.fit2$n.trees, 
    grid.resolution = 100,
    ice = TRUE
    ) %>%
  autoplot(rug = TRUE, train = bgr_train, alpha = .1)

#Look at Longitude
##average marginal effect on mpd.obs.z
gbm.fit2 %>%
  pdp::partial(pred.var = "long", n.trees = gbm.fit2$n.trees, grid.resolution = 100) %>%
  autoplot(rug = TRUE, train = bgr_train) +
  ylab("MPD.Z")

#ICE plot
gbm.fit2 %>%
  partial(
    pred.var = "long", 
    n.trees = gbm.fit2$n.trees, 
    grid.resolution = 100,
    ice = TRUE
    ) %>%
  autoplot(rug = TRUE, train = bgr_train, alpha = .1) 

#Look at NPP
##average marginal effect on mpd.obs.z
gbm.fit2 %>%
  pdp::partial(pred.var = "NPP", n.trees = gbm.fit2$n.trees, grid.resolution = 100) %>%
  autoplot(rug = TRUE, train = bgr_train) +
  ylab("MPD.Z")

#ICE plot
gbm.fit2 %>%
  partial(
    pred.var = "NPP", 
    n.trees = gbm.fit2$n.trees, 
    grid.resolution = 100,
    ice = TRUE
    ) %>%
  autoplot(rug = TRUE, train = bgr_train, alpha = .1) 
```


#How well do the boosted regession model predictions match the data?
```{r plot_bgr_predictions, echo = TRUE}
#predicting using the best model

# predict values for test data
pred <- predict(gbm.fit2, n.trees = gbm.fit2$n.trees, bgr_test)

# results
caret::RMSE(pred, bgr_test$mpd.obs.z) #[1] 0.08386642 close to RMSE using the best training model
```

#Figure 2
```{r fig2, echo = TRUE}
#pull dependecy data
gbm.fit2 %>%
  pdp::partial(pred.var = "log_leafshape", n.trees = gbm.fit2$n.trees, grid.resolution = 100)->ls
gbm.fit2 %>%
  pdp::partial(pred.var = "log_treeheight", n.trees = gbm.fit2$n.trees, grid.resolution = 100)->th
gbm.fit2 %>%
  pdp::partial(pred.var = "lat", n.trees = gbm.fit2$n.trees, grid.resolution = 100)->la
gbm.fit2 %>%
  pdp::partial(pred.var = "long", n.trees = gbm.fit2$n.trees, grid.resolution = 100)->lo
gbm.fit2 %>%
  pdp::partial(pred.var = "log_ntaxa", n.trees = gbm.fit2$n.trees, grid.resolution = 100)->sr
gbm.fit2 %>%
  pdp::partial(pred.var = "NPP", n.trees = gbm.fit2$n.trees, grid.resolution = 100)->np

#save datasets for making Figure 2
saveRDS(ls, "leafshape_brt.rds")
saveRDS(th, "treeheight_brt.rds")
saveRDS(la, "lat_brt.rds")
saveRDS(lo, "long_brt.rds")
saveRDS(sr, "speciesrichness_brt.rds")
saveRDS(np, "npp_brt.rds")

#leafshape
leaf_pdp<-ggplot()+
  geom_line(ls, mapping = aes(x= log_leafshape, y = yhat), size = 1)+
  geom_hline(yintercept = 0, linetype="dashed")+
  theme_classic()+
  labs(y= "Standard Effect Size of MPD", x = "log (Leaf Shape)")+
  scale_y_continuous(breaks= c(0, -1, -2, -3)) +
  theme(text = element_text(size=12),
        axis.title = element_text(size =12),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10))

#treeheight
height_pdp<-ggplot()+
  geom_line(th, mapping = aes(x= log_treeheight, y = yhat), size = 1)+
  geom_hline(yintercept = 0, linetype="dashed")+
  theme_classic()+
  labs(y = " ", x = "log (Tree Height)")+
  scale_y_continuous(breaks= c(0, -1, -2, -3)) +
  theme(text = element_text(size=12),
        axis.title = element_text(size =12),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10))

#latitude
lat_pdp<-ggplot()+
  geom_line(la, mapping = aes(x= lat, y = yhat), size = 1)+
  geom_hline(yintercept = 0, linetype="dashed")+
  theme_classic()+
  labs(y = " ", x = "Latitude")+
  scale_y_continuous(breaks= c(0, -1, -2, -3)) +
  theme(text = element_text(size=12),
        axis.title = element_text(size =12),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10))

#longitude
long_pdp<-ggplot()+
  geom_line(lo, mapping = aes(x= long, y = yhat), size = 1)+
  geom_hline(yintercept = 0, linetype="dashed")+
  theme_classic()+
  labs(y = " ", x = "Longitude")+
  scale_y_continuous(breaks= c(0, -1, -2, -3)) +
  theme(text = element_text(size=12),
        axis.title = element_text(size =12),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10))

#npp
npp_pdp<-ggplot()+
  geom_line(np, mapping = aes(x= NPP, y = yhat), size = 1)+
  geom_hline(yintercept = 0, linetype="dashed")+
  theme_classic()+
  labs(y = " ", x = "Net Primary Productivity")+
  scale_y_continuous(breaks= c(0, -1, -2, -3)) +
  theme(text = element_text(size=12),
        axis.title = element_text(size =12),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10))

#species richness
sr_pdp<-ggplot()+
  geom_line(sr, mapping = aes(x= log_ntaxa, y = yhat), size = 1)+
  geom_hline(yintercept = 0, linetype="dashed")+
  theme_classic()+
  labs(y = " ",  x = "log(Species Richness)")+
  scale_y_continuous(breaks= c(0, -1, -2, -3)) +
  theme(text = element_text(size=12),
        axis.title = element_text(size =12),
        axis.text.y = element_text(color="black", size = 10),
        axis.text.x = element_text(color="black", size = 10))

#title
title <- ggplot(data.frame(l = leaf_pdp$labels$y, x = 1, y = 1)) +
      geom_text(aes(x, y, label = l), angle = 90) + 
      theme_void() +
      coord_cartesian(clip = "off")

leaf_pdp$labels$y <- " "

library(patchwork)
comb<- (leaf_pdp + height_pdp + sr_pdp) / (lat_pdp + long_pdp + npp_pdp )
l_combo<-comb + plot_layout(guides = "collect") + plot_annotation(tag_levels = 'A')
title + l_combo + plot_layout(widths = c(1, 25))
```

```{r}
#Rescale functions
mm_rescale <- function(data, comparison){
  (data - min(data))/diff(range(data))*diff(range(comparison)) + min(comparison)
}

mm_rescale_getpoint <- function(data, comparison, point){
  my_perc <- (point - min(comparison))/diff(range(comparison))
  diff(range(data))*my_perc + min(data)
}

widen <- function(data, span){
  mean(range(data)) + c(-0.5, 0.5) *diff(range(data))*span
}

myplot <- function(mydf, xaxis, xlab = "", 
                   ylab = "Standard Effect Size of MPD", ylab2 = "Density",
                   myrange, xbreaks = NULL,
                   mybreaks = -3:0, secbreaks = NULL){
    ggplot()+
    geom_histogram(data = bgr_data2, 
                   mapping=aes(x=get(xaxis), y = ..density..), fill = "grey80")+
    geom_line(mydf, mapping = aes(x= get(xaxis), 
                                  y = mm_rescale(yhat, myrange)), size = 1)+
    geom_hline(yintercept = mm_rescale_getpoint(myrange, 
                                                mydf$yhat, 0),
               linetype="dashed")+
    theme_classic()+
    scale_x_continuous(name = xlab,
                       breaks = if(!is.null(xbreaks)){xbreaks} else{waiver()}) +
    scale_y_continuous(name =  ylab,
                       breaks = mm_rescale_getpoint(myrange, 
                                                    mydf$yhat, mybreaks),
                       labels = mybreaks,
                       sec.axis = sec_axis(trans =~.,
                                           name = ylab2,
                                           breaks = if(!is.null(secbreaks)){secbreaks} else{waiver()},
                                           labels = if(!is.null(secbreaks)){secbreaks} else{waiver()}
                                           )) +
    coord_cartesian(ylim = widen(mm_rescale_getpoint(myrange, mydf$yhat, mybreaks), 1.3)) +
    theme(text = element_text(size=12),
          axis.title = element_text(size =12),
          axis.text.y = element_text(color="black", size = 10),
          axis.text.x = element_text(color="black", size = 10))

}


myplot2 <- function(mydf, xaxis, xlab = "", 
                   ylab = "Standard Effect Size of MPD", ylab2 = "Density",
                   myrange, xbreaks = NULL,
                   mybreaks = -3:0, secbreaks = NULL){
    ggplot()+
    geom_histogram(data = bgr_data2, 
                   mapping=aes(x=get(xaxis), y = ..density..), fill = "grey80", binwidth = 1)+
    geom_line(mydf, mapping = aes(x= get(xaxis), 
                                  y = mm_rescale(yhat, myrange)), size = 1)+
    geom_hline(yintercept = mm_rescale_getpoint(myrange, 
                                                mydf$yhat, 0),
               linetype="dashed")+
    theme_classic()+
    scale_x_continuous(name = xlab,
                       breaks = if(!is.null(xbreaks)){xbreaks} else{waiver()}) +
    scale_y_continuous(name =  ylab,
                       breaks = mm_rescale_getpoint(myrange, 
                                                    mydf$yhat, mybreaks),
                       labels = mybreaks,
                       sec.axis = sec_axis(trans =~.,
                                           name = ylab2,
                                           breaks = if(!is.null(secbreaks)){secbreaks} else{waiver()},
                                           labels = if(!is.null(secbreaks)){secbreaks} else{waiver()}
                                           )) +
    coord_cartesian(ylim = widen(mm_rescale_getpoint(myrange, mydf$yhat, mybreaks), 1.3)) +
    theme(text = element_text(size=12),
          axis.title = element_text(size =12),
          axis.text.y = element_text(color="black", size = 10),
          axis.text.x = element_text(color="black", size = 10))
}


leaf_pdp <- myplot(ls, "log_leafshape", "log(Leaf Shape)", ylab = "", ylab2 = "", seq(0, 2, 0.2))
height_pdp <- myplot(th, "log_treeheight", "log(Tree Height)", ylab = "", ylab2 = "", seq(.6, 1.1, 0.2))
lat_pdp <- myplot2(la, "lat", "Latitude", ylab = "", ylab2 = "", seq(.02, .1, .01))
long_pdp <- myplot2(lo, "long", "Longitude", ylab = "", ylab2 = "", seq(0.03 , .05, 0.001))
sr_pdp <- myplot(sr, "log_ntaxa", "log(Species Richness)", ylab = "", ylab2 = "", 
                 xbreaks = seq(0, 6, 2), seq(0, 0.5, .001))
npp_pdp <- myplot(np, "NPP", "Net Primary Productivity", ylab = "", ylab2 = "", seq(.0063, .0077, .00001))

#title left
title <- ggplot(data.frame(l = "Standard Effect Size of MPD", x = .5, y = 1)) +
  geom_text(aes(x, y, label = l), angle = 90) + 
  theme_void() +
  coord_cartesian(clip = "off")

leaf_pdp$labels$y <- " "

#title right
title_right <- ggplot(data.frame(l = "Density", x = .5, y = 1)) +
  geom_text(aes(x, y, label = l), angle = 270) + 
  theme_void() +
  coord_cartesian(clip = "off")

leaf_pdp$labels$y2 <- " "

comb<- (leaf_pdp + height_pdp + sr_pdp) / (lat_pdp + long_pdp + npp_pdp )
l_combo<-comb + plot_layout(guides = "collect") + plot_annotation(tag_levels = 'A')

l_combo #add labels in keynote

ggsave("figure2.png", width = 12, height = 8)
```

