---
title: "Clean and collect Southern Africa tree data"
author: "APB"
date: "4/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(sf)
library(caper)
library(picante)
library(raster)
library(scico)
library(spdep)
library(gbm)
library(caret)
```

```{r load_data, include = FALSE}
tree<-read.nexus("data/36runs_BI250000_12K_output.tre")
#plot(tree, cex = 0.03) #1852 tips
data<-read.csv("data/SA_tree_data.csv", header = TRUE)

summary(data)
```

```{r clean_data, include =FALSE}
#first pull Family name out of the tree tip labels
tree$tip.label<-gsub("^.*?_","", tree$tip.label)

#check similarity between tree tips and data
tree_data<-data[which(data$matrix_name %in% tree$tip.label),]
tree_data$height<-as.numeric(tree_data$height)
tree_data$length<-as.numeric(tree_data$length)
tree_data$width<-as.numeric(tree_data$width)
tree_data$leaf_area<-as.numeric(tree_data$leaf_area)
summary(tree_data)

#remove any blanks from map_name
filter(tree_data, map_name != "")->tree_data

saveRDS(tree_data, "tree_data.rds")

length(unique(tree_data$map_name)) #there is data for 1085 map_names from the tree 

#subset the tree to only include taxa there is data for 
SA_tree<-drop.tip(tree, tree$tip.label[which(!(tree$tip.label %in% tree_data$matrix_name))])
#plot(SA_tree, cex = 0.3, type = "f")

saveRDS(SA_tree, "SA_tree.rds")
```

```{r calculate PD, include = FALSE}
#interspecific distance matrix
distance_matrix<-cophenetic.phylo(SA_tree)
```


```{r load_shape_files, include = FALSE}
#See some info on plotting shape files here:
#https://r-spatial.github.io/sf/articles/sf5.html
SA_poly<-st_read("data/GIS_DATA/all_no_exotics.shp")
plot(st_geometry(SA_poly)) #Contains 1176 polygons

SA_grids<-st_read("data/GIS_DATA/SA_grid.shp")
plot(st_geometry(SA_grids))

z<-subset(SA_poly, SA_poly$shpname=="acacia_karroo.shp")
plot(st_geometry(SA_grids))
plot(st_geometry(z), col = "lightblue", add = T) #add = T is adding the polygon for z to the previous plot

# outline of countries in Southern Africa
use<-map_data("world", region = c("Angola","Botswana","Lesotho","Mozambique","Malawi:3","Namibia", "Swaziland", "South Africa", "Zambia","Zimbabwe")) %>%
  dplyr::select(lon = long, lat = lat, group, id= region)

#remove island
use<-filter(use, lat > -40)
#remove extra countries 
use<-filter(use, lat < -15)

ggplot(use) +
  geom_polygon(mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.5)
```
#TO DO
## 1) overlay the grid and generate a simple species richness map
## 2) extract the list of species overlapping each grid cell and then calculate PD for each cell. 
## 3) map the median species trait values for each cell

###Notes: One way of doing all this is by ‘rasterising’ each species polygon, and then stacking them. Assign a ‘trait value’ to each raster (for species richness this would just be ‘1’), and then sum the values or take their mean etc. Some hints here: https://luisdva.github.io/rstats/richness/

#Species Richness Map
```{r, species_richness_map, echo = TRUE}
#QDS is quarter degree square grids
st_crs(SA_grids)<-4326
st_crs(SA_poly)
st_crs(SA_grids)

sf_use_s2(FALSE) # fixes the error with spherical geometry 

richness_grid <- SA_grids %>%
  st_join(SA_poly) %>%
  mutate(overlap = ifelse(!is.na(shpname), 1, 0)) %>% #1 if there is a shpname
  group_by(lat, long) %>%
  summarize(num_species = sum(overlap), 
            sp_names = list(shpname)) %>%
  mutate(ID = paste(lat, "_", long))

#dark grey/blue #2F4051

richness<-ggplot(richness_grid) +
  #geom_sf(data = SA_poly, fill = NA, size = 0.1) +
  geom_sf(aes(fill = num_species), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "Species richness") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.5)+
  theme_classic()+
  theme(
    #plot.background = element_rect(fill = "white"),
    #panel.background = element_rect(fill = "white"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    rect = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```

#Calculate PD for each grid cell (ID)
```{r list_of_species_in_each_grid_cell, include = FALSE}
#### ses.mpd from 'picante'
#need a community matrix where ID are the rows and species names are the columns
comm_matrix<-data.frame(matrix(ncol= length(tree$tip.label)), nrow=unique(richness_grid$ID), row.names = unique(richness_grid$ID))
rownames(comm_matrix)<-unique(richness_grid$ID)
tree_names<-tree_data[match(tree$tip.label, tree_data$matrix_name, nomatch = NA),] #put tree_data in tree tip.label order so we can easily switch between map_name and matrix_name
tree_names<-subset(tree_names, map_name != "NA")
colnames(comm_matrix)<-tree_names$map_name #set column names to the map names to make the for loop simpler

for (i in 1:nrow(comm_matrix)){
 comm_matrix[i,] <- colnames(comm_matrix) %in% unlist(subset(richness_grid, ID == rownames(comm_matrix)[i])$sp_names) %>% as.numeric

}

str(comm_matrix)
k<-as.numeric(comm_matrix['-34.75 _ 19.75',]) # should have a sum of 101
sum(k) # it does!

#make column names match the tree tip labels 
colnames(comm_matrix)<-tree_names$matrix_name

#matrix has extra columns that are unnecessary
comm_matrix<-comm_matrix[,1:1085]

##code takes time to run
#ses_results <-ses.mpd(comm_matrix, distance_matrix, null.model = "taxa.labels", runs = 1000) #30 minutes
##save output as RDS
#saveRDS(ses_results, "NRI_data.rds")
##read in ses.mpd results
NRI<-readRDS("NRI_data.rds") 
NRI$ID<-rownames(NRI)

#median
median(NRI$mpd.obs.z)

#24 grids have more dispersed species than we would expect by chance
dispersed<- subset(NRI, mpd.obs.z > 0) 
#1183 grids are clustered
clustered<-subset(NRI, mpd.obs.z < 0)

#species are more clustered than we would expect
plot(NRI$mpd.obs.z)
abline(h=0)
abline(h = -2.35, lty =2, col = "red") # median SES
```

#Map the mpd values for each cell
```{r mpd_plot, echo = TRUE}
right_join(richness_grid, NRI, by = "ID")->mpd_grid

#MPD map
MPD_plot<-ggplot(mpd_grid) +
  #geom_sf(data = SA_poly, fill = NA, size = 0.3) +
  geom_sf(aes(fill = mpd.obs), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "MPD") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), 
               fill = "transparent", 
               colour = "black", 
               size = 0.5)+
  theme_classic()+
  theme(panel.grid = element_blank(),
    legend.position = "right",
    rect = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank())

# SES map (more or less phylogenetically clustered than expected)
SESplot<-ggplot(mpd_grid) +
  #geom_sf(data = SA_poly, fill = NA, size = 0.3) +
  geom_sf(aes(fill = mpd.obs.z), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "SES") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.5)+
  theme_classic()+
  theme(panel.grid = element_blank(),
    legend.position = "bottom",
    rect = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank())

ggplot(mpd_grid) +
  #geom_sf(data = SA_poly, fill = "white", color = "grey", size = 0.05) +
  geom_sf(data = SA_grids, fill = "white", color = "black", size = 0.4) +
  geom_sf(aes(fill = mpd.obs.p), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "quantiles of SES") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.4)+
  theme_bw()+
  theme(
    #plot.background = element_rect(fill = "#f1f2f3"),
    #panel.background = element_rect(fill = "#2F4051"),
    #panel.grid = element_blank(),
    legend.position = "bottom",
    #line = element_blank(),
    #rect = element_blank()
    ) + labs(fill = "quantiles of SES (> 0.95 indicates phylogenetic evenness)")
# high quantiles (mpd.obs.p > 0.95) indicate phylogenetic evenness, or a greater phylogenetic distance among co-occurring species than expected.
```
#Summarise tree traits for each grid cell 
```{r pull_medians_for_trait_data, include = FALSE}
#dataframe by ID that includes the median tree trait data for each ID 
trait_data<-data.frame(ID = unique(richness_grid$ID), 
                       count =NA, 
                       treeheight = NA,
                       leafshape = NA,
                       leafarea = NA)

for (i in 1:nrow(trait_data)) {
  
  subset(tree_names, comm_matrix[i, ] == 1) ->j
  
  j %>%
   summarise(length(height)) %>% unlist() -> trait_data$count[i]
  
  j %>%
   summarise(mean(height, na.rm = T)) %>% unlist() -> trait_data$treeheight[i]
  
  j %>%
   summarise(mean(ratio, na.rm = T)) %>% unlist() -> trait_data$leafshape[i]
   
  j %>%
    summarise(mean(leaf_area, na.rm = T)) %>% unlist() -> trait_data$leafarea[i]
}

```

#Plot median tree traits per grid cell
```{r tree_trait_plots, echo =TRUE}
right_join(mpd_grid, trait_data, by = "ID")->trait_grid

height_plot<-ggplot(trait_grid) +
  #geom_sf(data = SA_poly, fill = NA, size = 0.3) +
  geom_sf(aes(fill = log(treeheight *1000)), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "Log (Tree Height in m)") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.5)+
  theme_classic()+
  theme(panel.grid = element_blank(),
    legend.position = "bottom",
    rect = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank())

shape_plot<-ggplot(trait_grid) +
  #geom_sf(data = SA_poly, fill = NA, size = 0.3) +
  geom_sf(aes(fill = leafshape), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "Leaf Shape") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.5)+
 theme_classic()+
theme(panel.grid = element_blank(),
    legend.position = "bottom",
    rect = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank())

ggplot(trait_grid) +
  #geom_sf(data = SA_poly, fill = NA, size = 0.3) +
  geom_sf(aes(fill = leafarea), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "Leaf Area") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.5)+
  theme_bw()+
  theme(legend.position = "bottom") + 
  labs(fill = "Leaf Area")

```

#Rerun JD's Stats
```{r correlations_regressions, include = TRUE}
#Is there a correlation between mpd and species richness? 0.33 ***
cor.test(mpd_grid$mpd.obs, mpd_grid$ntaxa)
cor.test(mpd_grid$mpd.obs.z, mpd_grid$ntaxa)

#Does mpd capture functional diversity? ...yes 59%
summary(lm(log(treeheight*1000) ~ mpd.obs * ntaxa, data =trait_grid))

#Is there a correlation between height and leaf shape? YES, tall trees have rounder leaves
cor.test(log(trait_grid$treeheight*1000), trait_grid$leafshape) # -0.75 ***

#Is there a ccorrelation between height and leaf area? Yes, tall trees have leaves with larger areas
cor.test(log(trait_grid$treeheight*1000), trait_grid$leafarea) #.817

#Is tree height explained by mpd and leaf shape?   65%
summary(lm(log(treeheight*1000) ~ mpd.obs + leafshape, data = trait_grid))

#Is tree height explained by mpd and leaf area?   67%
summary(lm(log(treeheight*1000) ~ mpd.obs + leafarea, data = trait_grid))
```

### need to pull AET data (and PET from some database) look for NPP net primary productivity data
#Adding NPP and AET data
```{r NPP_from_1996, include=FALSE, eval=FALSE}
#### NPP (kg-C/m2/year) from https://sage.nelson.wisc.edu/data-and-models/atlas-of-the-biosphere/mapping-the-biosphere/ecosystems#/net-primary-productivity/  ####
#Foley, J.A., I.C. Prentice, N. Ramankutty, S. Levis, D. Pollard, S. Sitch, and A. Haxeltine (1996) An Integrated Biosphere Model of Land Surface Processes, Terrestrial Carbon Balance and Vegetation Dynamics, Global Biogeochemical Cycles, 10, 603-628.
NPP_UWM<-raster("data/NPP_AET_data/NPP/npp/dblbnd.adf") #raster file
plot(NPP_UWM) 
crs(NPP_UWM) <- CRS('+init=EPSG:4326') #trying to add coordinates to the file

points<- extract(NPP_UWM, SA_grids)
str(points)
unlist(points)->b

#make a new dataframe with the NPP data
trait_npp_grid<-trait_grid
trait_npp_grid$npp<-b  #1196 grids of data

#plot the NPP data
ggplot(trait_npp_grid) +
  #geom_sf(data = SA_poly, fill = NA, size = 0.3) +
  geom_sf(aes(fill = npp), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "NPP (kg-C/m2/year)") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.3)+
  theme_bw()+
  theme(legend.position = "bottom")+ 
    labs(fill = "NPP")

#remove grids with NA for npp
trait_npp_clean<-subset(trait_npp_grid, npp != "NA")
```

```{r NPP_data_from_2013, include =FALSE}
##### Global Fire Emissions Database, Version 4.1 (GFEDv4) source: https://daac.ornl.gov/VEGETATION/guides/fire_emissions_v4_R1.html  ####
#trying to visualize the data
library(terra)
#f <- "~/Documents/UBC/SA_trees/NPP_AET_data/GFED4.1s_2013.hdf5"
#s <- sds(f) #S4 
#pull the data we want into a dataframe
#smaller<-data.frame(c(NPP = s$NPP, lat = s$lat, long = s$lon, grid_cell_area =s$grid_cell_area))
#subset the data to the lat and lon we are interested in
#subset(smaller, lat < -14) -> evenless
#subset(evenless, lon > 12) ->less
#subset(less, lon <37) ->less2
#subset(less2, lat > -35) -> less3

#make it an sf
#my.sf.point <- st_as_sf(x = less3, 
#                        coords = c("lon", "lat"),
#                        crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

#st_join(trait_npp_grid, my.sf.point)->test1
#saveRDS(test1, "trait_npp_data.rds")
##read npp data
trait_npp_data<-readRDS("trait_npp_data.rds") 

npp_plot<-ggplot(trait_npp_data) +
  #geom_sf(data = SA_poly, fill = NA, size = 0.3) +
  geom_sf(aes(fill = NPP), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "NPP (g C/m2/month)") +
  geom_polygon(use, mapping=aes(lon, lat, group = group), fill = "transparent", colour = "black", size = 0.5)+
theme_classic()+
theme(panel.grid = element_blank(),
    legend.position = "bottom",
    rect = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank())

```


All the figures together
```{r figure1, echo=FALSE}
all<-SESplot + npp_plot + richness + shape_plot + height_plot +  plot_layout(ncol = 3)

all  + plot_annotation(tag_levels = 'A')

ggsave("figure1.png", height = 6, width = 12)
```

#Check the ecotone region in the cape, it may explain why there is low species richness and high mpd



